Novel lattice Boltzmann method for simulation of strongly shear thinning viscoelastic fluids

### 2024-02-27 NovelLBM.tex
Updated comments

For a simple shear flow in the $xy$-plane of a PTT fluid with $ \xi =0$, an analytical expression for $ \eta $ and $N_1$ can be derived as detailed in appendix~\ref{app:semianalyticalPoiseuille}:
	\begin{equation}
		\eta\left(\dot{\gamma}\right) = \frac{\eta_\text{p}}{\exp\left[0.5W_0\left(4\epsilon \mathrm{Wi}^2\right)\right]} + \eta_\text{s} \label{eq:shearEta}
	\end{equation}
	\begin{equation}
		N_1 = \frac{\eta_\text{p}}{2\epsilon\lambda}W_0\left(4\epsilon \mathrm{Wi}^2\right) \label{eq:shearN}
	\end{equation}
where $\mathrm{Wi} = \lambda\dot{\gamma}$ is the Weissenberg number and $W_0$ is the Lambert $W$-function. \SG{ check that $ \eta _ \mathrm{s}$ is correct }\RK{what do you mean by correct?}
Fitting these equations to the rheological data for alginate and methyl cellulose leads to satisfactory agreement as shown in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
In the fit, the solvent viscosity was fixed at $\eta_\text{s} = \SI{1}{\milli\pascal\second}$ as the viscosity of the PBS buffer. 
%\SG{ give ref }\RK{PBS is 99\% water. If you really want a citation tho we could use doi: 10.1063/1.4823586} \SG{ ok, let's leave it out. }
%At very high shear rates, the contained polymers are completely stretched and the viscosity asymptotically approaches the solvent value $ \eta _ \mathrm{s}$.
The obtained parameters are given in table~\ref{tab:PTTparameters}.
	\begin{table}[H]
		\centering
		\begin{tabular}{c|c|c|c|c}
			Fluid & $\eta_\text{p}/\unit{\milli\pascal\second}$ & $\lambda/\unit{\milli\second}$ & $\epsilon$ & $R_\eta$ \\
			\hline
			alginate \SI{4}{\percent} & \num{48.2\pm0.4e3} & \num{343 \pm 4} & \num{0.545 \pm 0.010} & \num{4.8e4} \\
			methyl c. \SI{0.49}{\percent} & \num{18.7\pm0.4} & \num{0.344\pm0.003} & \num{0.270\pm0.003} & \num{18.7} \\
			methyl c. \SI{0.59}{\percent} & \num{32.5\pm0.7} & \num{0.433\pm0.004} & \num{0.365\pm0.004} & \num{32.5} \\
			methyl c. \SI{0.83}{\percent} & \num{81\pm2} & \num{0.714\pm0.009} & \num{0.496\pm0.006} & \num{81}
		\end{tabular}
		\caption{Parameters of the PTT model obtained by fitting to the rheological measurements in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
		\SG{ add a column with $R_ \eta $ }}\RK{would be identical to the content of the $\eta_p$ column, also $R_\eta$ would be used before introduced} \SG{ Stimmt, aber ich will es auf einen Blick sichtbar machen }
		\label{tab:PTTparameters}
	\end{table}

@@

\SG{ make sure that all papers cited in the introduction as ''3D'' are discussed here as well }\RK{dellar is missing}
\SG{ and vice versa: make sure that all papers discussed here are also cited in the introduction }\RK{gupta2015\_2 not in intro} \SG{ dann setz einfach die Zitate so rein, wie du sie fuer passend haeltst. }\todo{do}
Since the early days of LBM, variants for viscoelastic fluids have been developed.
Keeping in mind our target geometries, we focus here on three-dimensional formulations.
The first approaches avoided an explicit modeling of a viscoelastic constitutive equation and simply modified the LB collision operator to obtain a basic viscoelastic fluid obeying Jeffery's model \cite{lallemand2003}.
More advanced approaches include the work of Onishi \textit{et al.} \cite{onishi2005, onishi2006} who considered a non-shear-thinning Oldroyd-B model using a microscopic Fokker-Planck equation for the polymer dumbbells.
The work of \cite{wang2020} solved the advection part of the model using a second LBM running in parallel to the Navier-Stokes LBM.
The same technique was used by \cite{malaspinas2010} and \cite{ma2020} for the Oldroyd-B and FENE-P models.
Later, hybrid schemes coupling an LB solver for the solvent to a finite-difference \cite{gupta2015_2, gupta2015, gupta2016} or finite-volume \cite{kuron2021} scheme for the polymer stress (or, alternatively, the polymer conformation) tensor were introduced.
To improve numerical stability, some authors propose the inclusion of an artificial diffusivity \cite{malaspinas2010,ma2020,wang2020,wang2021}, which introduces a small error compared to the analytical solution of the constitutive equation.
Other remedies include special decompositions of the stress tensor\cite{dzanic2022} or optimization of the solvent contribution\cite{gupta2015} \SG{ rephrase and give some more details }\RK{I have not the slightest clue what gupta is doing. I assume optimize is code for increase until stable but I do not know.}.
Coupling of the resulting polymer stress to the Navier-Stokes LBM can be achieved via the inclusion of point forces or via a modification of the equilibrium populations.
A recent comparison in 2D found that the former in general leads to better stability \cite{dzanic2022_2}.\RK{No in fact, it uses the latter and found to be more accurate than a paper using the former, stability was entirely attributed to the decomposition and artificial diffusivity.} \SG{ die Referenz war falsch, es war der andere Dzanic. Aber da steht eigentlich schon, dass forcing besser sei, oder? }\todo{check/do not mention this}

@@

To include the polymer stress into the LBM, we use the stress coupling scheme according to Dzanic \textit{et al.} \cite{dzanic2022_2, gupta2015, onishi2006}.
For this, ... \SG{ do we really do stress coupling? Above it said that force coupling is better? }\RK{yes, no idea why you said that}
\SG{ give some more details, especially since we do SRT }.\RK{what does this have to do with SRT?}
\SG{ also include inflow/outflow BCs }\RK{done}
\SG{ do you advect polymer stress or polymer conformation? }\RK{A, which is most similar to polymer conformation}
To handle inflow/outflow boundaries, the polymer conformation tensor is specified on the boundary and advected using the corner transport upwind scheme.

@@

The key difficulty for simulating many realistic viscoelastic fluids is the imbalance between the viscous stress of the solvent and the polymer stress.
While the former is handled efficiently and reliably by the LBM streaming-collision scheme, the latter enters the LB equations as an additional source term.
The relative magnitude of the two terms is given by the viscosity ratio $R_\eta$ as defined in equation~\ref{eq:R_eta} which for many situations is of the order of $10^2-10^3$ (see figures~\ref{fig:fit_alginate}, \ref{tab:bioinks} and table~\ref{tab:bioinks}).
The key idea of our scheme is to artificially increase the LBM viscous stress while at the same time reducing the polymer stress by the same amount.
Starting from equation~\ref{eq:totalStress}, we write
\begin{align}
	\ma{\sigma} = 2\eta_\text{s}\ma{D} + \ma{\tau}_\text{shuffle} + \ma{\tau} - \ma{\tau}_\text{shuffle}	\label{eq:shuffleStress}
\end{align}
with the transferred ("shuffled") stress
\begin{align}
	\ma{\tau}_\text{shuffle} = 2\alpha_\text{s}\eta_\text{p}\ma{D}
\end{align}
containing an arbitrary parameter $\alpha_\text{s}$. \SG{ is there an allowed range for $ \alpha _ \mathrm{s}$? }\RK{LBM viscosity should stay positive; While $\alpha_s > 1$ stops decreasing the polymer stress and increases it again it can still be sensible as the relation of the stresses will still drop (which is desirable)}
Finally, this results in
\begin{align}
	\ma{\sigma} = 2\left(\eta_\text{s} + \alpha_\text{s}\eta_\text{p}\right)\ma{D} + \left(\ma{\tau} - 2\alpha_\text{s}\eta_\text{p}\ma{D}\right).
\end{align}
In practice, the shuffling parameter $\alpha_\text{s}$ is chosen before the start of the simulation.
The LBM solver is run with $\eta_\text{s} + \alpha_\text{s}\eta_\text{p}$ as its viscosity.
The polymer stress determined by equation~\ref{eq:polymerStress} is reduced by the amount $\alpha_\text{s}\eta_\text{p}\ma{D}$ at every time step before it is coupled into the LBM.

@@

We start by simulating the alginate and \SI{0.49}{\percent} methyl cellulose solutions with a viscosity ratio $R_\eta = \num{48.2e3}$ and $R_\eta = \num{18.7}$, respectively, in 2D planar Poiseuille flow.
In physical units the channel has a height of \SI{20}{\micro\meter} which are discretized with 41 LBM nodes.
For the viscosity shuffling, we set $\alpha_\text{s} = 1$.
Along the flow direction, the channel is periodic, and the flow is driven by a pressure gradient of \SI{-1e7}{\pascal\per\meter} for alginate and  approximately \SI{-1.2e7}{\pascal\per\meter} for methyl cellulose.
The velocity field is initialized with zero.
For methyl cellulose, we simulate the full temporal evolution with the polymers starting in the unstretched configuration.
In the case of alginate, we speed up the convergence and start from a situation in which the polymers are pre-stretched \SG{ give details }\RK{initialized to the semianalytical solution} \SG{ schreib am besten eine Referenz auf die Gleichung hin, muesste ja im Appendix stehen. }\todo{do}.
After \num{2e9} LBM time steps for alginate and \num{5e6} LBM time steps for methyl cellulose corresponding to approximately \SI{1.6}{\milli\second} and \SI{2.4}{\second} of physical time, respectively, the simulation is stopped and the flow field $u(y)$ is compared to the semi-analytical solution.
%\RK{
%	Alg 2e9 1.6455647956111251 ms (reveals the sim to be a scam)
%	MC 5e6 10.065719379854297 ms
%}
As shown in figure~\ref{fig:2Dvel}, very good agreement is found.
For a quantitative assessment, we compute the L2 error as follows.
\begin{equation}
	e_\text{L2} = \sqrt{\frac{\sum_i\left[\vec{u_i} - \vec{u}\left(y_i\right)\right]^2}{\sum_i\vec{u}\left(y_i\right)^2}}
\end{equation}

The spatially resolved errors are shown in figure~\ref{fig:2DvelErr} in the appendix.
Averaging over the channel diameter, we obtain an L2 error of \num{9e-4} and \num{1e-3}\RK{we do not "average" over the diameter, we calculate the L2 error along the diameter} for alginate and methyl cellulose, respectively.
These conclusions remain robust even when varying the pressure gradient, and thus the flow velocity, over many orders of magnitude as shown in figure~\ref{fig:pressureErr}~(a) of the appendix.\RK{This is the worst case error for MC and it is tiny "remain robust" is too weak here}

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/2DAlginateU.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/2DParameterStudyUWorst.pdf}
		\caption{(a) Velocity profile of alginate in planar channel flow.
		The time evolution of the center-line velocity is shown in the inset \ref{fig:2DTimeEvolution}\SG{ to do }\RK{Not an inset but plotted. Has to be inset?}.
		(b) Velocity profile for methyl cellulose.
		\SG{ make these plots for different $ \alpha _ \mathrm{s}$, either include here or put into appendix }\RK{Why? Is likely identical down to fp32 precision (see appendix C)}}
		\label{fig:2Dvel}
	\end{figure}

We proceed to evaluate the viscosity shuffling method for a cylindrical Poiseuille flow.
The semi-analytical solution for this situation can be derived similarly as for the planar case and is given in appendix~\ref{app:semianalyticalPoiseuille}.
Numerically, the cylindrical situation is slightly more complex due to discretization errors ("staircase effect") when mapping the rounded channel walls onto the rectilinear LBM grid \cite{Lehmann.20229da}.
We use the same material parameters as above, but double the pressure gradient to \SI{-2e7}{\pascal\per\meter} and approximately \SI{-2.5e7}{\pascal\per\meter} respectively. \SG{ why? } \RK{where analytical solutions exist in Poiseuille geometry the 3D solution differs from the 2D solution by a factor 2 (Due to the gradient of the stress in the derivation having a form in cylinder coordinates that can be solved by a factor 2). See the velocities in our sims being the same with this scaling}
The channel has a diameter of \SI{20}{\micro\meter} and is discretized with 41 lattice points.
Again, we initialize the polymers in the unstretched state for methyl cellulose and in the stretched state for alginate.
The resulting velocity profile can be seen in figure~\ref{fig:3Dvel} and shows almost equally good agreement as for the planar case with an L2 error of \num{0.014} and \num{0.06}, respectively.
The spatially resolved error in figure~\ref{fig:3DvelErr} of the appendix shows that this larger average is indeed mainly attributable to the staircase effect as the error near the walls increases significantly.
The robustness against pressure gradient changes is shown in figure~\ref{fig:pressureErr}~(b) of the appendix.

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/3DAlginateU.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/3DParameterStudyUWorst.pdf}
		\caption{(a) Velocity profile of alginate in cylindrical channel flow.
		The time evolution of the center-line velocity when starting from a pre-stretched configuration (see main text) is shown in the inset \ref{fig:3DTimeEvolution}.\RK{currently not an inset}
		(b) Velocity profile for methyl cellulose.
				\SG{ make these plots for different $ \alpha _ \mathrm{s}$, either include or put into Si }\RK{Why? Is likely identical down to fp32 precision (see appendix C)}}
		\label{fig:3Dvel}
	\end{figure}

%Malaspinas: We use richard/master fe0fc04115b9b6ea6628bcc6cb6593d3f72ac0ea (2023-11-05)
Finally, we assess the quality of our algorithm by comparing to an existing study on viscoelastic LBM by Malapsinas \textit{et al.} \cite{malaspinas2010}.
The authors studied an Oldroyd-B fluid in a planar Poiseuille flow and compared the error in the simulated velocity profile to an exact analytical solution.
The Oldroyd-B model is a special case of the PTT model which is obtained by setting $\xi = \epsilon = 0$ in eq.~(\ref{eq:polymerStress}) and is thus easily tractable with our code.\RK{was simulated with a direct Oldroyd-B implementation, which I now see was unnecessary...} \SG{ hast du mal kurz geprueft, ob dasselbe rauskommt? }\RK{Nein}\todo{check}
In figure~\ref{fig:malaspinas2010}, we show the L2 error as function of lattice resolution for different values of $\mathrm{Wi}$ and $R_ \nu $ which is defined as \cite{malaspinas2010}
\begin{align}
	R_\nu = \frac{\nu_\text{s}}{\nu_\text{p}+\nu_\text{s}}
\end{align}
with the kinematic solvent and the polymer viscosity $\nu_\text{s}$ and $\nu_\text{p}$. \SG{ comment whether these definitions are identical or different than the definitions above. }\RK{different but can be transformed into each other $\frac{1}{R_\nu}-1=R_\eta$}
Furthermore, in agreement with \cite{malaspinas2010}, also the L2 error is defined differently than above, namely:
	\begin{equation}\label{eq:l2error}
		E_\text{u} = \frac{1}{u_\text{max}}\sqrt{\frac{1}{N}\sum\abs{u_\text{simulation}-u_\text{theory}}^2}.
	\end{equation}
At this point, we mention that there is a typographic mistake in eq.~(61) of \cite{malaspinas2010} where the prefactor should be 8 instead of 4. \SG{ stimmt das so? }\RK{Ja}

@@

As can be seen in figure~\ref{fig:malaspinas2010}, the accuracy of our method does not depend on the Weissenberg Number and is generally somewhat better than the method of \cite{malaspinas2010}.
We note a certain dependence of the accuracy on $R_\nu$ which we consider nevertheless acceptable given the overall smallness of the error.
Finally, we find that the error of our method decreases with lattice resolution to the power of \num{-2} \SG{ give number, show dashed line in plot }\RK{done}.\RK{They see diminishing returns we don't}

@@

Our next example is the RT-DC geometry in which cells from a reservoir are squeezed through a narrow observation channel \cite{Toepfner.2018, Fregin:2019gx} with a square cross-section of 20x20 \unit{\micro\meter}.
For this simulation we use periodic boundary conditions and a constant body force corresponding to a pressure gradient of \SI{1e7}{\pascal}.
In figure~\ref{fig:RT-DC}, we show the velocity profile and the shear component of the stress which is the relevant component if a local shear viscosity is to be defined.
There are three qualitative differences compared to a purely viscous fluid.
Directly after the inlet we find a region of low stress.
Due to the fast advection, the polymers, despite being inside the constriction, have not reacted to the increased fluid stress yet.
Similarly, directly after the constriction, the high stress inside the observation channel persists over a certain distance that is needed for the polymer stress to decay.
\RK{There are three, proceeds to name two}

 	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/RT-DC_v.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/RT-DC_sigma_vM.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/RT-DC_tau.pdf}
		\caption{Velocity profile (a), von Mises stress (b) and shear stress (c) for methyl cellulose in a typical RT-DC geometry. For the calculated viscosity see figure~\ref{fig:Nozzle/RT-DC_eta}.\SG{ draw channel contour with black lines }\RK{done something else instead. Better?}
		\SG{ change to von Mises instead of $ \tau _{ 12 }$? }\RK{done something}}
		\label{fig:RT-DC}
	\end{figure}

@@

In the von Mises stress shown in figure \ref{fig:cellInShear_vM} the effect of the cell is clearly visible.
We observe four distinct regions of high stress radiating outwards from the cell body.
These are particularly pronounced along the longest axis of the deformed cell.
\SG{ connect to the ''viscoelastic wings'' paper \cite{Zhang.2020}}\RK{Our "wings" look very different}

@@

\section{Semi-analytical solution for Poiseuille flows of PTT fluids with solvent viscosity}
\label{app:semianalyticalPoiseuille}
\SG{ make a self-contained derivation which is understandable even without reading Oliveira \\}\RK{done}
\SG{ use Gl.~(\ref{eq:polymerStress}) as starting point and end with Poiseuille solution, but also shear solution Gl.~(\ref{eq:shearN}, \ref{eq:shearEta}) \\}\RK{done}
Our derivation is in principal similar to Oliveira \cite{oliveira1999}. For both Poiseuille flow and shear flow the following holds.
	\vec{u} = \dot{\gamma}\left(y\right)\hat{e}_x
With this the constitutive equation~(\ref{eq:polymerStress}) for $\xi=0$ in steady state reduces to the following.

@@

With $\Tr\ma{\tau} = \tau_{xy}$. In the three-dimensional case $y$ refers to the radial coordinate.

>>>>

For a simple shear flow in the $xy$-plane of a PTT fluid with $ \xi =0$, an analytical expression for $ \eta $ and $N_1$ can be derived as detailed in appendix~\ref{app:semianalyticalPoiseuille}:
	\begin{equation}
		\eta\left(\dot{\gamma}\right) = \frac{\eta_\text{p}}{\exp\left[0.5W_0\left(4\epsilon \mathrm{Wi}^2\right)\right]} + \eta_\text{s} \label{eq:shearEta}
	\end{equation}
	\begin{equation}
		N_1 = \frac{\eta_\text{p}}{2\epsilon\lambda}W_0\left(4\epsilon \mathrm{Wi}^2\right) \label{eq:shearN}
	\end{equation}
where $\mathrm{Wi} = \lambda\dot{\gamma}$ is the Weissenberg number and $W_0$ is the Lambert $W$-function.
Fitting these equations to the rheological data for alginate and methyl cellulose leads to satisfactory agreement as shown in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
In the fit, the solvent viscosity was fixed at $\eta_\text{s} = \SI{1}{\milli\pascal\second}$ as the viscosity of the PBS buffer. 
%\SG{ give ref }\RK{PBS is 99\% water. If you really want a citation tho we could use doi: 10.1063/1.4823586} \SG{ ok, let's leave it out. }
%At very high shear rates, the contained polymers are completely stretched and the viscosity asymptotically approaches the solvent value $ \eta _ \mathrm{s}$.
The obtained parameters are given in table~\ref{tab:PTTparameters}.
	\begin{table}[H]
		\centering
		\begin{tabular}{c|c|c|c|c}
			Fluid & $\eta_\text{p}/\unit{\milli\pascal\second}$ & $\lambda/\unit{\milli\second}$ & $\epsilon$ & $R_\eta$ \\
			\hline
			alginate \SI{4}{\percent} & \num{48.2\pm0.4e3} & \num{343 \pm 4} & \num{0.545 \pm 0.010} & \num{4.8e4} \\
			methyl c. \SI{0.49}{\percent} & \num{18.7\pm0.4} & \num{0.344\pm0.003} & \num{0.270\pm0.003} & \num{18.7} \\
			methyl c. \SI{0.59}{\percent} & \num{32.5\pm0.7} & \num{0.433\pm0.004} & \num{0.365\pm0.004} & \num{32.5} \\
			methyl c. \SI{0.83}{\percent} & \num{81\pm2} & \num{0.714\pm0.009} & \num{0.496\pm0.006} & \num{81}
		\end{tabular}
		\caption{Parameters of the PTT model obtained by fitting to the rheological measurements in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
		\SG{ add a column with $R_ \eta $ }}\RK{would be identical to the content of the $\eta_p$ column, also $R_\eta$ would be used before introduced} \SG{ Stimmt, aber ich will es auf einen Blick sichtbar machen }
		\label{tab:PTTparameters}
	\end{table}

@@

\SG{ make sure that all papers cited in the introduction as ''3D'' are discussed here as well }\RK{dellar is missing}
\SG{ and vice versa: make sure that all papers discussed here are also cited in the introduction }\RK{gupta2015\_2 not in intro} \SG{ dann setz einfach die Zitate so rein, wie du sie fuer passend haeltst. }\todo{do}
Since the early days of LBM, variants for viscoelastic fluids have been developed.
Keeping in mind our target geometries, we focus here on three-dimensional formulations.
The first approaches avoided an explicit modeling of a viscoelastic constitutive equation and simply modified the LB collision operator to obtain a basic viscoelastic fluid obeying Jeffery's model \cite{lallemand2003}.
More advanced approaches include the work of Onishi \textit{et al.} \cite{onishi2005, onishi2006} who considered a non-shear-thinning Oldroyd-B model using a microscopic Fokker-Planck equation for the polymer dumbbells.
The work of \cite{wang2020} solved the advection part of the model using a second LBM running in parallel to the Navier-Stokes LBM.
The same technique was used by \cite{malaspinas2010} and \cite{ma2020} for the Oldroyd-B and FENE-P models.
Later, hybrid schemes coupling an LB solver for the solvent to a finite-difference \cite{gupta2015_2, gupta2015, gupta2016} or finite-volume \cite{kuron2021} scheme for the polymer stress (or, alternatively, the polymer conformation) tensor were introduced.
To improve numerical stability, some authors propose the inclusion of an artificial diffusivity \cite{malaspinas2010,ma2020,wang2020,wang2021}, which introduces a small error compared to the analytical solution of the constitutive equation.
Other remedies include special decompositions of the stress tensor\cite{dzanic2022} or optimization of the solvent contribution\cite{gupta2015} \SG{ rephrase and give some more details }\RK{I have not the slightest clue what gupta is doing. I assume optimize is code for increase until stable but I do not know.}.
Coupling of the resulting polymer stress to the Navier-Stokes LBM can be achieved via the inclusion of point forces or via a modification of the equilibrium populations.
Both methods have recently been compared in 2D \cite{dzanic2022_2}.

@@

To include the polymer stress into the LBM, we use the stress coupling scheme according to Dzanic \textit{et al.} \cite{dzanic2022_2, gupta2015, onishi2006}.
For this, ... \SG{ more details }
To handle inflow/outflow boundaries, the polymer conformation tensor is specified on the boundary and advected using the corner transport upwind scheme.

@@

The key difficulty for simulating many realistic viscoelastic fluids is the imbalance between the viscous stress of the solvent and the polymer stress.
While the former is handled efficiently and reliably by the LBM streaming-collision scheme, the latter enters the LB equations as an additional source term.
The relative magnitude of the two terms is given by the viscosity ratio $R_\eta$ as defined in equation~\ref{eq:R_eta} which for many situations is of the order of $10^2-10^3$ (see figures~\ref{fig:fit_alginate}, \ref{tab:bioinks} and table~\ref{tab:bioinks}).
The key idea of our scheme is to artificially increase the LBM viscous stress while at the same time reducing the polymer stress by the same amount.
Starting from equation~\ref{eq:totalStress}, we write
\begin{align}
	\ma{\sigma} = 2\eta_\text{s}\ma{D} + \ma{\tau}_\text{shuffle} + \ma{\tau} - \ma{\tau}_\text{shuffle}	\label{eq:shuffleStress}
\end{align}
with the transferred ("shuffled") stress
\begin{align}
	\ma{\tau}_\text{shuffle} = 2\alpha_\text{s}\eta_\text{p}\ma{D}
\end{align}
containing an arbitrary parameter $\alpha_\text{s}$. 
Finally, this results in
\begin{align}
	\ma{\sigma} = 2\left(\eta_\text{s} + \alpha_\text{s}\eta_\text{p}\right)\ma{D} + \left(\ma{\tau} - 2\alpha_\text{s}\eta_\text{p}\ma{D}\right).
\end{align}
In practice, the shuffling parameter $\alpha_\text{s}$ is chosen before the start of the simulation.
The LBM solver is run with $\eta_\text{s} + \alpha_\text{s}\eta_\text{p}$ as its viscosity.
The polymer stress determined by equation~\ref{eq:polymerStress} is reduced by the amount $\alpha_\text{s}\eta_\text{p}\ma{D}$ at every time step before it is coupled into the LBM.

@@

We start by simulating the alginate and \SI{0.49}{\percent} methyl cellulose solutions with a viscosity ratio $R_\eta = \num{48.2e3}$ and $R_\eta = \num{18.7}$, respectively, in 2D planar Poiseuille flow.
In physical units the channel has a height of \SI{20}{\micro\meter} which are discretized with 41 LBM nodes.
For the viscosity shuffling, we set $\alpha_\text{s} = 1$.
Along the flow direction, the channel is periodic, and the flow is driven by a pressure gradient of \SI{-1e7}{\pascal\per\meter} for alginate and  approximately \SI{-1.2e7}{\pascal\per\meter} for methyl cellulose.
The velocity field is initialized with zero.
For methyl cellulose, we simulate the full temporal evolution with the polymers starting in the unstretched configuration.
In the case of alginate, we speed up the convergence and start from a situation in which the polymers are pre-stretched \SG{ give details }\RK{initialized to the semianalytical solution} \SG{ schreib am besten eine Referenz auf die Gleichung hin, muesste ja im Appendix stehen. }\todo{do}.
After \num{2e9} LBM time steps for alginate and \num{5e6} LBM time steps for methyl cellulose corresponding to approximately \SI{1.6}{\milli\second} and \SI{2.4}{\second} of physical time, respectively, the simulation is stopped and the flow field $u(y)$ is compared to the semi-analytical solution.
%\RK{
%	Alg 2e9 1.6455647956111251 ms (reveals the sim to be a scam)
%	MC 5e6 10.065719379854297 ms
%}
As shown in figure~\ref{fig:2Dvel}, very good agreement is found.
For a quantitative assessment, we compute the L2 error as follows.
\begin{equation}
	e_\text{L2} = \sqrt{\frac{\sum_i\left[\vec{u_i} - \vec{u}\left(y_i\right)\right]^2}{\sum_i\vec{u}\left(y_i\right)^2}}
\end{equation}
which gives \num{9e-4} and \num{1e-3} for alginate and methyl cellulose, respectively.
The spatially resolved errors are shown in figure~\ref{fig:2DvelErr} in the appendix.
These conclusions remain unchanged even when varying the pressure gradient, and thus the flow velocity, over many orders of magnitude as shown in figure~\ref{fig:pressureErr}~(a) of the appendix.

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/2DAlginateU.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/2DParameterStudyUWorst.pdf}
		\caption{(a) Velocity profile of alginate in planar channel flow.
		The time evolution of the center-line velocity is shown in the inset \ref{fig:2DTimeEvolution}\SG{ to do }\RK{Not an inset but plotted. Has to be inset?}.
		(b) Velocity profile for methyl cellulose.
}
		\label{fig:2Dvel}
	\end{figure}

We proceed to evaluate the viscosity shuffling method for a cylindrical Poiseuille flow.
The semi-analytical solution for this situation can be derived similarly as for the planar case and is given in appendix~\ref{app:semianalyticalPoiseuille}.
Numerically, the cylindrical situation is slightly more complex due to discretization errors ("staircase effect") when mapping the rounded channel walls onto the rectilinear LBM grid \cite{Lehmann.20229da}.
We use the same material parameters as above, but double the pressure gradient to \SI{-2e7}{\pascal\per\meter} and approximately \SI{-2.5e7}{\pascal\per\meter}, respectively, thereby mirroring the relation in Newtonian Poiseuille flow where twice the pressure is required to obtain the same flow rate in cylindrical compared to planar gemeotries.
The channel has a diameter of \SI{20}{\micro\meter} and is discretized with 41 lattice points.
Again, we initialize the polymers in the unstretched state for methyl cellulose and in the stretched state for alginate.
The resulting velocity profile can be seen in figure~\ref{fig:3Dvel} and shows almost equally good agreement as for the planar case with an L2 error of \num{0.014} and \num{0.06}, respectively.
The spatially resolved error in figure~\ref{fig:3DvelErr} of the appendix shows that this larger average is indeed mainly attributable to the staircase effect as the error near the walls increases significantly.
The robustness against pressure gradient changes is shown in figure~\ref{fig:pressureErr}~(b) of the appendix.

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/3DAlginateU.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/3DParameterStudyUWorst.pdf}
		\caption{(a) Velocity profile of alginate in cylindrical channel flow.
		The time evolution of the center-line velocity when starting from a pre-stretched configuration (see main text) is shown in the inset \ref{fig:3DTimeEvolution}.\RK{currently not an inset}
		(b) Velocity profile for methyl cellulose.
				}
		\label{fig:3Dvel}
	\end{figure}

%Malaspinas: We use richard/master fe0fc04115b9b6ea6628bcc6cb6593d3f72ac0ea (2023-11-05)
Finally, we assess the quality of our algorithm by comparing to an existing study on viscoelastic LBM by Malapsinas \textit{et al.} \cite{malaspinas2010}.
The authors studied an Oldroyd-B fluid in a planar Poiseuille flow and compared the error in the simulated velocity profile to an exact analytical solution.
The Oldroyd-B model is a special case of the PTT model which is obtained by setting $\xi = \epsilon = 0$ in eq.~(\ref{eq:polymerStress}) and is thus easily tractable with our code.\RK{was simulated with a direct Oldroyd-B implementation, which I now see was unnecessary...} \SG{ hast du mal kurz geprueft, ob dasselbe rauskommt? }\RK{Nein}\todo{check}
In figure~\ref{fig:malaspinas2010}, we show the L2 error as function of lattice resolution for different values of $\mathrm{Wi}$ and $R_ \nu $ which is defined as \cite{malaspinas2010}
\begin{align}
	R_\nu = \frac{\nu_\text{s}}{\nu_\text{p}+\nu_\text{s}}
\end{align}
with the kinematic solvent and the polymer viscosity $\nu_\text{s}$ and $\nu_\text{p}$. 
We note that this can directly be transformed into the viscosity ratio used above as $\frac{1}{R_\nu}-1=R_\eta$.
Furthermore, in agreement with \cite{malaspinas2010}, also the L2 error is defined differently than above, namely:
	\begin{equation}\label{eq:l2error}
		E_\text{u} = \frac{1}{u_\text{max}}\sqrt{\frac{1}{N}\sum\abs{u_\text{simulation}-u_\text{theory}}^2}.
	\end{equation}
At this point, we mention that there is a typographic mistake in eq.~(61) of \cite{malaspinas2010} where the prefactor should be 8 instead of 4. \SG{ stimmt das so? }\RK{Ja}

@@

As can be seen in figure~\ref{fig:malaspinas2010}, the accuracy of our method does not depend on the Weissenberg Number and is generally somewhat better than the method of \cite{malaspinas2010}.
We note a certain dependence of the accuracy on $R_\nu$ which we consider nevertheless acceptable given the overall smallness of the error.
Finally, we find that the error of our method decreases with lattice resolution to the power of \num{-2}. \RK{They see diminishing returns we don't} \SG{ explain }

@@

Our next example is the RT-DC geometry in which cells from a reservoir are squeezed through a narrow observation channel \cite{Toepfner.2018, Fregin:2019gx} with a square cross-section of 20x20 \unit{\micro\meter}.
For this simulation we use periodic boundary conditions and a constant body force corresponding to a pressure gradient of \SI{1e7}{\pascal}.
In figure~\ref{fig:RT-DC}, we show the velocity profile and the shear component of the stress which is the relevant component if a local shear viscosity is to be defined.
There are three qualitative differences compared to a purely viscous fluid.
Directly after the inlet we find a region of low stress.
Due to the fast advection, the polymers, despite being inside the constriction, have not reacted to the increased fluid stress yet.
Similarly, directly after the constriction, the high stress inside the observation channel persists over a certain distance that is needed for the polymer stress to decay.
\RK{There are three, proceeds to name two} \SG{ ? }

 	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/RT-DC_v.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/RT-DC_sigma_vM.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/RT-DC_tau.pdf}
		\caption{Velocity profile (a), von Mises stress (b) and shear stress (c) for methyl cellulose in a typical RT-DC geometry. For the calculated viscosity see figure~\ref{fig:Nozzle/RT-DC_eta}.
		\SG{ change to von Mises instead of $ \tau _{ 12 }$? }\RK{done something}}
		\label{fig:RT-DC}
	\end{figure}

@@

In the von Mises stress shown in figure \ref{fig:cellInShear_vM} the effect of the cell is clearly visible.
We observe four distinct regions of high stress radiating outwards from the cell body.
These are particularly pronounced along the longest axis of the deformed cell.

@@

\section{Semi-analytical solution for Poiseuille flows of PTT fluids with solvent viscosity}
\label{app:semianalyticalPoiseuille}
Our derivation is to the one in Oliveira \cite{oliveira1999}. For both Poiseuille flow and shear flow, we assume that the velocity is unidirectional: 
\begin{equation}
	\vec{u} = \dot{\gamma}\left(y\right)\hat{e}_x.
\end{equation}
\SG{ This equation does not work, $ \dot{ \gamma  }$ is a shear rate, it should read $ \partial u_x / \partial y = \dot{ \gamma  }$. }

With this the constitutive equation~(\ref{eq:polymerStress}) for $\xi=0$ in steady state reduces to

@@

for a planar flow.
In axisymmetric flows $y$ in eqs.~(\ref{eq:PTTDerivation1}, \ref{eq:PTTDerivation2}) needs to be replaced by the radial coordinate $r$. \SG{ no factor 2? }
With $\Tr\ma{\tau} = \tau_{xy}$. \SG{ should be $ \tau _{ xx }$? }
